Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-21 _tex/paper2.tex
Built site for gh-pages

Although replacing the Normal with the Logistic distribution has a
minimal impact on trait estimation or model interpretation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}, Thurstone
acknowledged that the two remaining assumptions of Case V--and, by
extension, of the BTL model--could pose significant problems when
researchers assess complex traits or heterogeneous stimuli
\citep{Thurstone_1927b}. In particular, he noted that while assuming
equal discriminal dispersions and zero correlation for discriminal
processes simplifies the trait measurement model, these assumptions may
fail to reflect the complexity of CJ data
\citep{Thurstone_1927a, Andrich_1978, Kelly_et_al_2022}. As a result,
they can lead to unreliable and inaccurate trait estimates
\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.

Furthermore, because Thurstone aimed primarily to produce a trait's
``coarse scaling'' and ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
how to use the resulting trait estimates for statistical inference. The
CJ tradition has attempted to address this gap by separating trait
estimation from the hypothesis testing process. Specifically, CJ studies
often rely on point estimates of the traits (i.e., the BTL scores)--or
their transformations--to conduct statistical inference. Although
separating these processes simplifies the analysis of CJ data, the
statistical literature cautions against this practice, as it can
introduce bias into the analysis and compromise the reliability of the
hypothesis test
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

@@

The explicit integration of the sampling and comparison mechanisms
within our conceptual model offers a novel perspective on the first
research avenue. Specifically, it shows how these mechanisms influence
the data-generating process by acting as missing data sources,
determining which observations are part of the data sample. This new
perspective invites the application of Little and Rubin's principled
missing data framework \citep{Little_et_al_2020}, enabling a more
rigorous evaluation of existing claims about these mechanisms, their
influence on CJ outcomes, and their implications for designing and
evaluating more complex experimental setups.

One prominent claim in the CJ literature is that Thurstone's model
naturally handles non-random missing data without compromising the
reliability or validity of trait estimates \citep{Bramley_2008}.
However, it is easy to imagine scenarios where this assumption may not
hold. For instance, selecting a particular group of judges could result
in an incomplete representation of the trait's full range and
complexity, as judgment accuracy varies with demographic factors
\citep{Bartholomew_et_al_2020a, Kelly_et_al_2022} and even between
individuals
\citep{Gill_et_al_2013, vanDaal_et_al_2016, vanDaal_et_al_2017, vanDaal_2020}.
Consequently, judges' selection could affect the formation of a ``shared
consensus'', potentially biasing trait estimates and distorting the
observed trait distribution \citep{Deffner_et_al_2022}. It is therefore
essential to determine whether such sampling biases occur--and if so,
whether they undermine the validity of trait estimates and the
robustness of statistical inferences. Our framework offers a rigorous
setting to test the scenario.

@@

Moreover, when designing and evaluating more complex experimental
setups, our framework highlights one key principle: the sampling and
comparison mechanisms directly influence the CJ outcome. In this study,
we deliberately structured these mechanisms to produce data that is
\emph{missing completely at random} (MCAR)--that is, the missingness is
unrelated to any observed or unobserved variables, including the outcome
\citep{Little_et_al_2020}. This design offers one key advantage: it
generates simple random samples that satisfy the condition of
ignorability, enabling researchers to legitimately ignore missing data
during analysis without introducing bias
\citep{Everitt_et_al_2010, Kohler_et_al_2019}.

However, many modern CJ applications involve more complex forms of
missingness--such as \emph{missing at random} (MAR) or \emph{missing not
at random} (MNAR) mechanism \citep{Little_et_al_2020}--that arise from
the chosen sampling and comparison methods. Two notable examples are
adaptive comparative judgment designs
\citep{Pollitt_2012a, Pollitt_2012b} and case-control studies
\citep{Everitt_et_al_2010}. In adaptive designs, an algorithm defines
the comparison mechanism by dynamically selecting which stimulus pairs
judges evaluate. These designs have produced mixed evidence: some
studies report improved trait reliability
\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, while others
suggest these gains may be artificially inflated
\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.
In contrast, case-control studies rely on sampling mechanisms linked to
non-assignable variables that likely influence the trait of interest.
For instance, \citet{Boonen_et_al_2020} selected children based on
hearing status, a non-assignable variable that likely affects speech
quality \citep{Boonen_et_al_2023, Rivera_et_al_2024}. Nevertheless,
regardless of their missingness mechanism, both designs could benefit
from explicitly articulating their underlying causal structures. Doing
so would enable researchers to assess and test their impact on trait
estimation and statistical inference. Our framework provides a rigorous
foundation for such articulations.

Additionally, by explicitly incorporating judges' biases, our framework
provides a novel approach to investigate the second research avenue.
Specifically, it enables researchers to determine whether such biases
occur and, if so, to assess their impact on the reliability and validity
of trait estimation. Although some anecdotal evidence already suggests
judges' biases exist
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a},
further research is needed to address several key questions. For
instance: Can CJ data be analyzed under the assumption of
``sample-free'' trait calibration--that is, the estimated trait
difference for any two scripts does not depend on which other scripts
they are compared with, implying an absence of judges' bias?
\citep{Bramley_2008, Andrich_1978} If this assumption does not hold, how
many judges and judgments are required to ensure acceptable levels of
trait reliability and validity? Do training and expertise influence
judgment biases, and to what extent? \citep{Kelly_et_al_2022} Finally,
what are the benefits of \emph{repeated measures designs}, in which
judges evaluate the same stimulus pairs multiple times?
\citep{Lawson_2015}

>>>>

Although replacing the Normal with the Logistic distribution has a
minimal impact on trait estimation or model interpretation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}, Thurstone
acknowledged that the two remaining assumptions of Case V--and, by
extension, of the BTL model--could pose significant problems when
researchers assess complex traits or heterogeneous stimuli
\citep{Thurstone_1927b}. Specifically, while assuming assuming equal
discriminal dispersions and zero correlation for discriminal processes
simplifies the trait measurement model, these assumptions may fail to
reflect the complexity of CJ data
\citep{Thurstone_1927a, Andrich_1978, Kelly_et_al_2022}. As a result,
they can lead to unreliable and inaccurate trait estimates
\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.

Furthermore, because Thurstone aimed primarily to produce a trait's
``coarse scaling'' and ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
how to use the resulting trait estimates for statistical inference. The
CJ tradition has attempted to address this gap by separating trait
estimation from hypothesis testing. Specifically, CJ studies often rely
on point estimates of the traits--typically the BTL scores or their
transformations--to conduct statistical inference. Although this
separation simplifies CJ data analysis, the statistical literature warns
that this practice can introduce bias into the analysis and compromise
the reliability of the hypothesis tests
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

@@

The explicit integration of the sampling and comparison mechanisms
within our conceptual model offers a novel perspective on the first
research avenue. Specifically, it shows how these mechanisms influence
the data-generating process by acting as missing data sources,
determining which observations are part of the data sample. This new
perspective invites the application of Little and Rubin's
\citeyearpar{Little_et_al_2020} principled missing data framework,
enabling a more rigorous evaluation of existing claims about these
mechanisms, their influence on CJ outcomes, and their implications for
designing and evaluating more complex experimental setups.

One prominent claim in the CJ literature is that Thurstone's model
naturally handles non-random missing data without compromising the
reliability or validity of trait estimates \citep{Bramley_2008}.
However, it is easy to imagine scenarios where this assumption may not
hold. For instance, selecting a particular group of judges could result
in an incomplete representation of the trait's full range and
complexity, as judgment accuracy varies with demographic factors
\citep{Bartholomew_et_al_2020a, Kelly_et_al_2022} and even between
individuals
\citep{Gill_et_al_2013, vanDaal_et_al_2016, vanDaal_et_al_2017, vanDaal_2020}.
Consequently, judges' selection could affect the formation of a ``shared
consensus'', potentially biasing trait estimates and distorting the
observed trait distribution \citep{Deffner_et_al_2022}. It is therefore
essential to determine whether such sampling biases undermine the
validity of trait estimates and the robustness of statistical
inferences. Our framework offers a rigorous setting to test this
scenario.

@@

Moreover, when designing and evaluating more complex experimental
setups, our framework highlights one key principle: the sampling and
comparison mechanisms directly influence the CJ outcome. In this study,
we deliberately structured these mechanisms to produce data that is
\emph{missing completely at random} (MCAR)--that is, the missingness is
unrelated to any observed or unobserved variables, including the outcome
\citep{Little_et_al_2020}. This design offers one key advantage: it
generates simple random samples that satisfy the condition of
ignorability, enabling researchers to legitimately \emph{ignore} missing
data during analysis without introducing bias
\citep{Everitt_et_al_2010, Kohler_et_al_2019}.

However, many modern CJ applications involve more complex forms of
missingness--such as \emph{missing at random} (MAR) or \emph{missing not
at random} (MNAR) mechanism \citep{Little_et_al_2020}--that arise from
the chosen sampling and comparison methods. Two notable examples are
adaptive comparative judgment designs
\citep{Pollitt_2012a, Pollitt_2012b} and case-control studies
\citep{Everitt_et_al_2010}. In adaptive designs, an algorithm defines
the comparison mechanism by dynamically selecting which stimulus pairs
judges evaluate. These designs have produced mixed evidence: some
studies report improved trait reliability
\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, while others
suggest these gains may be artificially inflated
\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.
In contrast, case-control studies rely on sampling mechanisms linked to
non-assignable variables that likely influence the trait of interest.
For instance, \citet{Boonen_et_al_2020} selected children based on
hearing status, a non-assignable variable that likely affects speech
quality \citep{Boonen_et_al_2023, Rivera_et_al_2024}. Nevertheless,
regardless of their missingness mechanism, both designs could benefit
from explicitly articulating their underlying causal structures. Doing
so would enable researchers to assess and test their impact on trait
estimation and statistical inference. Our framework, again, provides a
rigorous foundation for such articulations.

Additionally, by explicitly incorporating judges' biases, our framework
provides a novel approach to investigate the second research avenue.
Specifically, it enables researchers to determine whether such biases
occur and, if so, to assess their impact on the reliability and validity
of trait estimation. Although some anecdotal evidence already suggests
judges' biases exist
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a},
further research is needed to address several critical questions. For
instance, can CJ data be analyzed under the assumption of
``sample-free'' trait calibration, which implies an absence of judges'
bias? \citep{Bramley_2008, Andrich_1978} If this assumption does not
hold, how many judges and judgments per judge are required to ensure
acceptable levels of trait reliability and validity? Do training and
expertise influence judgment biases, and to what extent?
\citep{Kelly_et_al_2022} Finally, what are the benefits of
\emph{repeated measures designs}, in which judges evaluate the same
stimulus pairs multiple times, in assessing judgment accuracy?
\citep{Lawson_2015}
